---
title: "Analysis of Well-Designed Experiment"
author: " "
date: "October 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Temperature and moisture impact strongly on the early stages of a plant’s life cycle. 
Here, we investigate how temperature and moisture affect these early stages of plant development in four Banksia species.

Six polycarbonate-covered steel exclusion shelters were constructed on site to manipulate water availability and assess effects on plant growth as measured by biomass at 6 weeks post-germination. Two garden beds were placed within each shed. Garden beds within a shed were randomised to receive either a warming soil treatment or no warming. Three watering treatments per shed: drought, normal watering, or over-watering were also evaluating within each shed.  

We predicted that reduced moisture and increased temperature would delay plant growth over the 6-week evaluation period.

In our data, we collected Shed number (1-6), Garden Bed (1-2) within each Shed, and watering Section (1-3) within each shed since these factors may contribute to the variation in biomass. 

Import libraries
```{r}

```


Import data. We re-level Water so that it is ordered by amount.
```{r}
banksia <- read_csv("Data/Banksia data.csv") %>%
  mutate(Water = factor(Water, levels = c("low","med","high")))

glimpse(banksia)
```


Summarise the data. How many sheds? How many plants per shed? Create a table to show that the data are balanced.
```{r}

```

Visualise the layout of temperature treatment across the sheds. 
```{r}

```

Visualise the layout of water treatment across the sheds. 
```{r}

```

Choose a shed and visualise the layout of the species within that shed. 
```{r}

```

Exploring biomass and the factors that impact on biomass.
Questions to explore:
How does temperature affect biomass? Is the temperature effect differ by species? by water treatment? 

How does water treatment affect biomass? Does this effect differ by species? by temperature?
```{r}

```

Fit a linear model to address the research question. Take data structure into account.
```{r}
lm1 <-
```

Interpret the ANOVA table
```{r}
anova(lm1)
```

Assess the residual plots. Is the model resonable?
```{r}

```

Use the summary function and look at the Random Effects. Which factors contribute the most to the variation in biomass?
```{r}

```

As an exercise, try removing one of the random factors from your model. Does it affect the inference?
```{r}

```


Use emmeans() to extract the mean biomass and SE estimates from the model for each species, water treatment and temperature. Create a graphic that summarises the results you wish to highlight.
```{r}

```

